Competing Constructions for German Measure Phrases

### 2017-03-18 Latex/article.tex
theory part almost done

However, if the measure noun denotes a container, constructions with \textit{voll} (\textit{von})\slash\textit{voller} `full (of)' and \textit{mit} `with' are available, see (\ref{ex:alternatives}).

@@

This section reviews some existing analyses of the PGC-NAC alternation or related issues, and I develop my own analysis and the appropriate hypotheses for the studies in Sections~\ref{sec:corpusstudies} and \ref{sec:externalvalidation}.
While the constructions and the alternation are often covered in descriptive grammars such as \citet[XYZ]{Eisenberg2013a}, the theoretical literature has little specific discussion to offer.

A caveat first:
I do \textit{not} review the semantics literature on partitives and pseudo-partitives because the constructions of interest in the present paper are all pseudo-partitives, and thus none of the variance in the choice of the PGC and NAC can be expected to be explained by a discussion along these lines.
The same is mostly true for the work on the grammaticalisation of pseudo-partitives and measure noun constructions, which has two main streams.
On one hand, it is often assumed that pseudo-partitives arise as a form of grammaticalised partitives (\eg\ \citealp[536--539]{Koptjevskaja2001} for Finnish and Estonian, \citealp[559]{Koptjevskaja2001} for European languages per se).%
\footnote{Notice that the description of German in \citet[547--549]{Koptjevskaja2001} is vague and partly incorrect, maybe because her main source \cite{Eschenbach1993} is not as clear as it could be (see also Footnote \ref{fn:eschenbash} above).
Her example (29b) from \citet[71]{Eschenbach1993}, repeated here verbatim as (\ref{ex:koptjewskaja}) is misclassified as a clear case of a dative on the measure noun and a genitive on the kind noun.

@@

It could, however, equally well be a genitive on the measure noun because of the syncretisms discussed here on p.\ \pageref{page:femininesyncretism} and an ambiguous case requirement of the preposition \textit{troz}.
Then, she does not clearly distinguish acceptable from non-acceptable constructions (cf.\ Section \ref{sec:descriptive}) on her pp.\ 548 and 549, including a rather confusing \textit{???} rating of her (31b), which is clearly a non-acceptable construction of the type illustrated in (\ref{ex:analyticalpartitive:b}) above.
}
Since the alternating constructions are both clearly pseudo-partitives and not in any way partitives, this diachronic development (now completed) is not expected to have any effect on the alternation.
On the other hand, there are more pertinent approaches like \cite{Brems2003,DeclerckBrems2016}.
In English -- \cite{Brems2003} argues based on a view from \cite{Langacker1991} -- strongly lexicalised and thus semantically bleached measure nouns turn from nouns into quantifiers, leading to a reanalysis of the NP, wherein the kind noun becomes the head.
This is supported by convincing syntactic evidence, for example subject-verb agreement.
In German MNPs, the measure noun always remains the syntactic head, however (see immediately below for references), and I know of no strong hypotheses stating that the degree of grammaticalisation of the measure noun influences the alternation in any way.
The statistical models presented in Section~\ref{sec:corpusstudies} could in principle show, however, whether highly grammaticalised measure nouns favor the NAC or the PGC through the estimation of lemma-specific random intercepts.

A major interest in traditional and generative linguistics is to find a plausible analysis of the hierarchical syntactic structure of the constructions in question.
In the case at hand, the first question is which noun constitutes the head of the whole construction.
This question was answered quite clearly by \cite{Loebel1986} already (see also \citealp[213]{Eschenbach1994} or \citealp[16]{GallmannLindauer1994}).
Subject-verb agreement is always realised on the measure noun, and we can therefore assume that the measure noun is the head.
Turning to the MNP-internal syntax, it cannot be understood properly without taking into account the strong\slash weak inflection patterns in German described in Section~\ref{sec:descriptive} (especially Footnote~\ref{fn:strongweak}).
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the so-called \textit{weak} pattern showing massive syncretism.
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners.
Thus, the adjectives in the target constructions have properties of adjectives (they are lexical adjectives) and determiners (inflection and leftmost position in the NP) in a sense.
In this light, the pattern shown in Table~\ref{tab:constructions} makes perfect sense, especially in a probabilistic framework:
If speakers use classify the adjective as a determiner, they have to use the PGC because with a determiner, the PGC is the only option.
If speakers classify the adjective as an adjective, the kind NP has no determiner, and the NAC is the only option.
While the generative analysis presented in \cite{Bhatt1990} cannot deal with such probabilistic effects, she comes quite close to this interpretation by analysing the kind NP in the GPC as a DP and in the NAC as an NP.

% Syntactic structure: adjectives-as-determiners and layers in NP, role assignment to nouns in PGC/NAC
%     Gallmann => Bhatt1990 DP[gen] vs NP[agree]
%     Mention problem of no case assigner for NAC, so sub-syntactic process or theta theory wrong

Finally, I collect some hypotheses (or maybe mostly observations) from the existing literature as to which factors influence the choice of NAC and PGC.

% Influencing factors: Hentschel, Zimmer
% Squeeze in Anttila & Fong

@@

% On Case-OCP (from Götz Keydana):
% Und da habe ich dann auch einen Verweis auf den locus classicus gefunden:
% STEFAN A. FRISCH, JANET B. PIERREHUMBERT and MICHAEL B. BROE: SIMILARITY AVOIDANCE AND THE OCP. Natural Language & Linguistic Theory 22: 179–228, 2004.
% Dazu gäbe es dann noch, ebenfalls von Pierrehumbert,
% Dissimilarity in the Arabic Verbal Roots. In Amy Schafer, ed.,Proceedings of NELS 23 Amherst, MA: Graduate Linguistics Student Association. 1993. pp 367-381 
% Darin stellt sie wohl erstmals explizit einen Zusammenahng zwischen OCP und Kognition her.

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{figures/corpus/04_glmm_raneff_mn}~\hspace{0.05\textwidth}~\includegraphics[width=0.5\textwidth]{figures/corpus/04_glmm_raneff_fem}
\caption{Sample random effects for the masculine\slash neuter and feminine GLMMs}
\label{fig:glmm:fixef:minus1pos}
\end{figure}

@@

>>>>

However, if the measure noun denotes a container or vessel, constructions with \textit{voll} (\textit{von})\slash\textit{voller} `full (of)' and \textit{mit} `with' are available, see (\ref{ex:alternatives}).

@@

This section reviews some existing analyses of the PGC-NAC alternation or related issues, and I develop my own analysis and the appropriate hypotheses for the studies in Sections~\ref{sec:corpusstudies} and \ref{sec:externalvalidation}.
While the constructions and the alternation are often covered in descriptive grammars such as \citet[XYZ]{Eisenberg2013a}, the available linguistic literature contains little specific discussion of the situation in German.
In The remainder of this section, I discuss as factors that might influence the alternation:
degrees of grammaticalisation of the measure noun, the different syntactic structures of the variants and their syntactic prototypes, and potential effects of semantic prototypicality.

Turning to grammaticalisation, it is often assumed that pseudo-partitives arise as a form of grammaticalised partitives (\eg\ \citealp[536--539]{Koptjevskaja2001} for Finnish and Estonian, \citealp[559]{Koptjevskaja2001} for European languages per se).%
\footnote{Notice that the description of German in \citet[547--549]{Koptjevskaja2001} is vague and partly incorrect, maybe because her main source \cite{Eschenbach1993} is not as clear as it could be (see also Footnote \ref{fn:eschenbash} above).
Her example (29b) from \citet[71]{Eschenbach1993}, repeated here verbatim as (\ref{ex:koptjewskaja}) is misclassified as a clear case of a dative on the measure noun and a genitive on the kind noun (hence, PGC).

@@

It could, however, equally well be a genitive on the measure noun because of the syncretisms discussed here on p.\ \pageref{page:femininesyncretism} and an ambiguous case requirement of the preposition \textit{troz}, hence an instance of NAC.
Then, she does not clearly distinguish acceptable from non-acceptable constructions (cf.\ Section \ref{sec:descriptive}) on her pp.\ 548 and 549, including a rather confusing \textit{???} rating of her (31b), which is clearly a non-acceptable construction of the type illustrated in (\ref{ex:analyticalpartitive:b}) above.
}
The alternating constructions in German are both clearly pseudo-partitives, but the grammaticalisation paths assumed by \cite[esp.\ 526--530]{Koptjevskaja2001}, which I interpret freely here, are clearly pertinent.
The development usually begins with constructions involving two referential nouns (not necessarily forming a single NP) and a \textit{separative} meaning as in \textit{(cut) two slices from the cake} \citep[535]{Koptjevskaja2001}.
The \textit{part-of} meaning of true partitives as in \textit{a slice of the case} represents the first stage of a development wherein the measure noun already tends to loose some semantic content.
The pseudo-partitive stage finally instantiates a \textit{quantity-of} relation, potentially even leading to fully grammaticalised quantifiers as in English \textit{a lot}.
Since in German, the PGC is clearly the older variant \citep{Zimmer2015} with the overt \textit{quantity-of} marker in the form of the genitive suffix and at least the potential to form a true partitive (if the kind noun is definite), which the NAC lacks, we can expect more strongly grammaticalised measure nouns to prefer the NAC.
Since clearly non-referential nouns like \textit{Gramm} `gram' and \textit{Meter} `metre' should favor the NAC according to this hypothesis, the effect can be detected by including information about measure noun classes in the corpus-based models in Section~\ref{sec:corpusstudies}.

On the other hand, there are case studies like \cite{Brems2003,DeclerckBrems2016} where syntactic evidence for grammaticalisation plays a prominent role.
In English -- \cite{Brems2003} argues based on a view from \cite{Langacker1991} -- strongly lexicalised and thus semantically bleached measure nouns turn from nouns into quantifiers, leading to a syntactic reanalysis of the NP, wherein the kind noun becomes the head.
The argument is supported by convincing syntactic evidence, for example subject-verb agreement.
In German MNPs, the measure noun always remains the syntactic head, however (see immediately below for references), and there is no similar clear criterion to detect grammaticalisation.
The following analysis of the morpho-syntax of German MNPs will however, provide hypothetical cues along similar lines.

A major interest in descriptive and generative linguistics is to find a plausible analysis of the hierarchical syntactic structure of the constructions in question.
In the case at hand, the first question is which noun constitutes the head of the whole construction.
This question was answered quite clearly by \cite{Loebel1986} already (see also \citealp[213]{Eschenbach1994}, and \citealp[16]{GallmannLindauer1994}).
As hinted above, subject-verb agreement is always realised on the measure noun, and we can therefore assume that the measure noun is the head.

I therefore turn to the MNP-internal syntax, which will show itself to be highly revealing.
It cannot be understood properly without taking into account the strong\slash weak inflection patterns of adjectives described in Section~\ref{sec:descriptive} (especially Footnote~\ref{fn:strongweak}).
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretistic \textit{weak} pattern.%
\footnote{The structural cases (nominative and accusative) in the singular are marked with \textit{-e}, the oblique cases (dative and genitive) as well as the entire plural are marked with \textit{-en}.
The only exception to this rule is the masculine accusative singular, which is also marked with \textit{-en}.}
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners, which is called the \textit{strong} pattern (see Table~\ref{tab:syncretisms}).
Thus, the adjectives in the target constructions have properties of adjectives as well as determiners in a sense:
they are lexical adjectives and function as attributive modifiers, but they are inflected as determiners and are the leftmost element in the NP.
This unusual double nature of adjectives in NPs without a determiner leads to a plausible interpretation of the pattern shown in Table~\ref{tab:constructions}.
Whenever speakers classify the adjective in the kind noun phrase as a determiner, they have to use the PGC because, if there is a determiner, the PGC is the only option.
When they classify the adjective as an adjective, the kind NP has no determiner, and the NAC is the only option.%
\footnote{While the generative analysis presented in \cite{Bhatt1990} cannot properly deal with probabilistic effects, Bhatt comes close to this interpretation by analysing the kind NP in the GPC as a DP and in the NAC as an NP.
Within the confines of her framework, this is the appropriate was of encoding the relevant generalisation.}
In a framework that assumes similarity-based probabilistic classification, such as Prototype Theory, we can now regard the PGC with a determiner in the kind noun phrase as instantiating the PGC prototype and the NAC with just a bare kind noun as the NAC prototype.
The choice of the alternating variants is then expected to be influenced by the degree of similarity to either of these prototypes.

The situation is made difficult by the fact that the semantic differences between the variants are at best subtle, which means the prototypes are not associated with nice and clearly detectable differences in meaning.
However, both the kind noun lemmas and the measure noun lemmas probably appear with different relative frequencies in the NAC with a bare kind noun and the PGC with a determiner inside the kind noun phrase (in part possibly due to their meaning and usage).
Their relative frequencies in these non-alternating configurations should be indicative of their preference to appear in one of the two variants.
In other words, lemmas themselves might have preferences for one prototype or the other.

%Related to the , dative measure nouns favor the PGC more than nominative and accusative (also \citealp{Hentschel1993,Zimmer2015}).
%As an example, \textit{mit einem Stück frischen Brots} `with a piece of fresh bread' (PGC) would be preferred over \textit{mit einem Stück frischem Brot} (NAC).
%I propose

Furthermore, the grammaticalisation path as described above leads from NPs denoting highly individuated objects standing in a \textit{separated-from} relation to a construction with a \textit{part-of} meaning (still available in the prototypical PGC) to a construction with a more diffuse \textit{quantity-of} relation (characteristic of the prototypical NAC).
The best evidence for the relevance of this factor in an ongoing grammaticalisation process is that plural kind nouns almost always trigger PGC and virtually never NAC (Section~\ref{sec:descriptive}).
More indirectly, plurality of the measure noun could favor the PGC over the NAC, and exact quantification in the form of cardinals even more so.%
\footnote{The effect of the plural was empirically shown but not explained in \cite{Hentschel1993}.}
As a highly illustrative example (using a measure noun which has virtually developed two variants, one of which grammaticalsed to a much higher degree), \textit{ein Haufen französische Watte} `a heap\slash lot of French cotton' (NAC) has essentially two uses.
It can be used to denote an unspecific amount of cotton in the sense of \textit{a lot of French cotton} in situations when the exact quantity of cotton is not of interest.
Under the appropriate circumstances, it could also be used to denote a highly specific quantity of cotton, for example when a cotton merchant always sells her French cotton in \textit{heaps} weighing two stone each.
The former is highly restricted for \textit{zwei Haufen französischer Watte} `two heaps/\textdagger lots of French cotton' (GPC).
Such an NP is less likely to be used in situations where the exact quantity of cotton is irrelevant, even if there are actually two heaps of cotton.
In other words, a precisely quantified measure noun leads to a higher individuation (the creation of portions) of the substance denoted by the kind noun, which is more appropriate for the PGC prototype.%
\footnote{In the most extreme, some measure nouns cannot be pluralised, such as \textit{eine Menge} `a lot of'.
Also note that the presentation of the examples in NAC and PGC is suggestive.
Native speakers will confirm that both NPs sound highly acceptable in both GPC and NAC.
It was made clear from the beginning and throughout that the preferences are indeed subtle.}

Finally, it is claimed that the PGC is more typical of higher registers or even exclusive to written language (see \citealp[320--323]{Hentschel1993}, where she also refers to descriptive grammars).
This is not surprising in as much as the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial and vernacular variants of German, and thus overrepresented to the same degree as colloquial, vernacular, and spoken variants are overrepresented in most corpora of written German.
There is no reason to assume that such preferences for cannot be part of the construction prototypes.
In the corpus study, the alternation tendencies controlled by this complex factor related to register, medium, and lectal variation will be tested using certain very simple proxies.

To summarise the discussion and core hypotheses, it is expected that:

\begin{enumerate}[i.]
  \item classes of more clearly grammaticalised measure nouns favor the NAC,
  \item the relative frequencies with which measure noun lemmas and kind noun lemmas appear in the prototypical (unequivocal) PGC and NAC are positively correlated with the relative frequency the PGC and the NAC are chosen in the alternating case,
  \item MNPs with masculine dative kind nouns disfavour the GPC comparatively less
  \item pluralic measure nouns and measure nouns quantified by cardinals favor the PGC variant
  \item .
\end{enumerate}

In the next section, I report the corpus study that was designed to test these hypotheses on usage data.

@@

@@

%\begin{figure}[h]
%\centering
%\includegraphics[width=0.5\textwidth]{figures/corpus/04_glmm_raneff_mn}~\hspace{0.05\textwidth}~\includegraphics[width=0.5\textwidth]{figures/corpus/04_glmm_raneff_fem}
%\caption{Sample random effects for the masculine\slash neuter and feminine GLMMs}
%\label{fig:glmm:fixef:minus1pos}
%\end{figure}

@@

%\footnote{\cite{AnttilaFong2000} propose an \textit{Obligatory Contour Principle for case} (Case-OCP) in a superficially similar but fundamentally unrelated case alternation in Finnish partitives.
%}
%
% On Case-OCP (from Götz Keydana):
% Und da habe ich dann auch einen Verweis auf den locus classicus gefunden:
% STEFAN A. FRISCH, JANET B. PIERREHUMBERT and MICHAEL B. BROE: SIMILARITY AVOIDANCE AND THE OCP. Natural Language & Linguistic Theory 22: 179–228, 2004.
% Dazu gäbe es dann noch, ebenfalls von Pierrehumbert,
% Dissimilarity in the Arabic Verbal Roots. In Amy Schafer, ed.,Proceedings of NELS 23 Amherst, MA: Graduate Linguistics Student Association. 1993. pp 367-381 
% Darin stellt sie wohl erstmals explizit einen Zusammenahng zwischen OCP und Kognition her.
